\Large The state learner -- a super learner for right-censored data

### 2024-12-21 manuscript/manuscript.tex
small changes

We define the state learner, a new super learner for right-censored
data, which simultaneously estimates the expected loss of learners of
the event time distribution and the censoring distribution. Compared
to the super learners of \cite{polley2011-sl-cens} and
\cite{golmakani2020super}, the state learner proposed here poses no
restrictions on the type of learners that can be included in the
library. Compared to super learners that rely on inverse probability
of censoring weighting, censoring unbiased transformations, or
pseudo-value based super learners, the state learner does not require
a pre-specified censoring distribution. Compared to the super learners
by \cite{han2021inverse} and \cite{westling2021inference}, the state
learner has theoretical guarantees which these methods do not
have.
% Moreover, state learner seems to be the first super learner for
% right censored data which learns the outcome and the censoring
% distribution simulaneously.

The loss function underlying the state learner operates on the
coarsened data. In the right-censored survival setting the coarsened
data consist of the minimum and the order of the censoring time and
the event time as well as the baseline covariates. The state learner
can include a broad class of survival models as learners, it can
handle competing risks, and it does not require a single pre-specified
estimator of the conditional censoring distribution. To analyse the
theoretical properties of the state learner we focus on the discrete
super learner which combines the library of learners by picking the
one that minimises the cross-validated loss \citep{van2007super}. In
the presence of competing risks, our algorithm uses separate libraries
of learners for the cumulative hazard functions for each of the
competing risks and for the censoring distribution. We show that the
oracle selector of the state learner is consistent if all libraries
contain a consistent learner and prove a finite sample oracle
inequality.

@@

There are many existing super learners for right-censored data. We briefly
review them here. Machine learning based on right-censored data commonly uses
the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with right-censored data
by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
time. However, the partial log-likelihood loss does not work well with data
splitting (cross-validation) in continuous time. The reason is that the partial
log-likelihood loss assigns an infinite value for any learner that predicts
piece-wise constant cumulative hazard functions when there are event times in
the test set which do not occur in the learning set. This problem occurs with
prominent survival learners including the Kaplan-Meier estimator, the random
survival forest, and the semi-parametric Cox regression model and these learners
cannot be included in the library of the super learner proposed by
\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
baseline hazard function can be profiled out of the likelihood
\citep{cox1972regression}. The cross-validated partial log-likelihood loss
\citep{verweij1993cross} has therefore been suggested as a loss function for
super learning by \cite{golmakani2020super}. This choice of loss function
restricts the library of learners to include only Cox proportional hazards
models, and hence excludes many learners such as, e.g., random survival models,
additive hazards models, and accelerated failure time models. Other proposals
for super learning with right-censored data use an inverse probability of
censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
methods rely on an estimator of the censoring distribution, and their drawback
is that this estimator has to be pre-specified. An approach which avoids a
pre-specified censoring model was proposed independently by
\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
authors suggest to iterate between learning of the outcome model and learning of
the censoring model using IPCW loss functions. However, it seems that this
procedure is in general not guaranteed to converge to the true data-generating
mechanism \citep{munch2024thesis}.

>>>>

We define the state learner, a new super learner for right-censored data, which
simultaneously estimates the expected loss of learners of the event time
distribution and the censoring distribution. Compared to the super learners of
\cite{polley2011-sl-cens} and \cite{golmakani2020super}, the state learner
proposed here poses no restrictions on the type of learners that can be included
in the library. Compared to super learners that rely on inverse probability of
censoring weighting, censoring unbiased transformations, or pseudo-values, the
state learner does not require a pre-specified censoring distribution. Compared
to the super learners by \cite{han2021inverse} and \cite{westling2021inference},
the state learner has theoretical guarantees which these methods do not have.
% Moreover, state learner seems to be the first super learner for
% right censored data which learns the outcome and the censoring
% distribution simulaneously.

The loss function underlying the state learner operates on coarsened data. In
the right-censored survival setting the coarsened data consist of the minimum
and the order of the censoring time and the event time as well as the baseline
covariates. The state learner works by simultaneously assessing how well
learners of the event time distribution and the censoring distribution predict
the states of the coarsened data. To analyse the theoretical properties of the
state learner we focus on the discrete super learner which combines the library
of learners by picking the one that minimises the cross-validated loss
\citep{van2007super}. In the presence of competing risks, our algorithm uses
separate libraries of learners for the cumulative hazard functions for each of
the competing risks and for the censoring distribution. We show that the oracle
selector of the state learner is consistent if all libraries contain a
consistent learner and prove a finite sample oracle inequality.

@@

There are many existing super learners for right-censored data. We briefly
review them here. Machine learning based on right-censored data commonly uses
the partial log-likelihood as a loss function
\citep[e.g.,][]{li2016regularized,yao2017deep,lee2018deephit,katzman2018deepsurv,gensheimer2019scalable,lee2021boosted,kvamme2021continuous}.
This loss function is also suggested for super learning with right-censored data
by \cite{polley2011-sl-cens}, where data is assumed to be observed in discrete
time. However, the partial log-likelihood loss does not work well with data
splitting (cross-validation) in continuous time. The reason is that the partial
log-likelihood loss assigns an infinite value for any learner that predicts
piece-wise constant cumulative hazard functions when there are event times in
the test set which do not occur in the learning set. This problem occurs with
prominent survival learners including the Kaplan-Meier estimator, the random
survival forest, and the semi-parametric Cox regression model and these learners
cannot be included in the library of the super learner proposed by
\cite{polley2011-sl-cens}. When a proportional hazards model is assumed, the
baseline hazard function can be profiled out of the likelihood
\citep{cox1972regression}. The cross-validated partial log-likelihood loss
\citep{verweij1993cross} has therefore been suggested as a loss function for
super learning by \cite{golmakani2020super}. This choice of loss function
restricts the library of learners to include only Cox proportional hazards
models, and hence excludes many learners such as, e.g., random survival forests,
additive hazards models, and accelerated failure time models. Other proposals
for super learning with right-censored data use an inverse probability of
censoring weighted (IPCW) loss function
\citep{graf1999assessment,van2003unicv,molinaro2004tree,keles2004asymptotically,hothorn2006survival,gerds2006consistent,gonzalez2021stacked},
censoring unbiased transformations
\citep{fan1996local,steingrimsson2019censoring}, or pseudo-values
\citep{andersen2003generalised,mogensen2013random,sachs2019ensemble}. All these
methods rely on an estimator of the censoring distribution, and their drawback
is that this estimator has to be pre-specified. An approach which avoids a
pre-specified censoring model was proposed independently by
\cite{han2021inverse} and \cite{westling2021inference}. In both articles, the
authors suggest to iterate between learning of the outcome model and learning of
the censoring model using IPCW loss functions. However, it seems that this
procedure is in general not guaranteed to converge to the true data-generating
mechanism \citep[][Appendix~A.4]{munch2024thesis}.
